---
output:
  html_document:
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(mongolite)
library(stringr)
library(Hmisc)
library(lubridate)
library(knitr)
library(kableExtra)

options(knitr.table.format = 'html')

# sites <- read_csv('sites.csv') %>%
#   select(siteID, region)
# 
# partRs <- read_csv('partRs.csv')

blitz <- read_csv('staticBlitz_processedRecords.csv') %>%
  mutate(date = as.Date(date))
# 
# siteProjectTable <- read_csv('siteProjectTable.csv')

summarize <- dplyr::summarize

```

<h1>The Neighborhood Nestwatch Resight Blitz<br> March 2018</h1>
<hr>

<p>Click <a href="https://raw.githubusercontent.com/bsevansunc/openData/master/staticBlitz_processedRecords.csv" target="_blank" download="blitz_processedRecords.csv">this link</a> to access the processed Blitz records.</p>

<!-- <br> -->
<!-- <a href="staticBlitz_processedRecords.csv" target="_blank">this link</a> -->
<br>
<h2>Blitz results, summarized across regions</h2>
<br>
<ul>
<li><b>Number of resights</b>:
<ul>
<li>Blitz: 252</li>
<li>Previous year, same period: 48 (525 % improvement!)</li>
<li>Previous week, same year: 76</li>
</ul>
</li>
<br>
<li><b>Number of resighted birds</b>:
<ul>
<li>Blitz: 145</li>
<li>Previous year, same period: 43 (337 % improvement!)</li>
<li>Previous week, same year: 68</li>
</ul>
</li>
<br>
<li><b>Number of participants that resighted birds</b>:
<ul>
<li>Blitz: 64</li>
<li>Previous year, same period: 29 (220 % improvement!)</li>
<li>Previous week, same year: 36</li>
</ul>
</li>
<br>
</ul>
<hr>
<h2>Blitz results, summarized by region</h2>
<br>
<div class = "row">
<div class = "col-sm-3 col-sm-offset-0">
<h3>Resights</h3>
</div>
<div class = "col-sm-1"></div>
<div class = "col-sm-3">
<h3>Resighted birds</h3>
</div>
<div class = "col-sm-1"></div>
<div class = "col-sm-4">
<h3>Participants that resighted</h3>
</div>
</div>


<div class = "row">
<div class = "col-sm-3 col-sm-offset-0">
```{r 'resights by region', echo = FALSE}
blitz %>%
  group_by(region) %>%
  summarize(n = n()) %>%
  kable('html') %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c('striped', 'hover', full_width = F)
  )
```
</div>
<div class = "col-sm-1"></div>
<div class = "col-sm-3">
```{r 'resighted birds by region', echo = FALSE}
blitz %>%
  select(-date) %>%
  distinct %>%
  group_by(region) %>%
  summarize(n = n()) %>%
  kable('html') %>%
  kable_styling(full_width = TRUE,
    bootstrap_options = c('striped', 'hover', full_width = F)
    )
```
</div>
<div class = "col-sm-1"></div>
<div class = "col-sm-3">
```{r 'number of participants that resighted birds', echo = FALSE, message = FALSE, warning = FALSE}
blitz %>%
  group_by(region) %>%
  summarize(n = length(unique(siteID))) %>%
  kable('html') %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c('striped', 'hover', full_width = F)
  )
```
</div>
<div class = "col-sm-1"></div>
</div>
<br>
<div class = "row">
<div class = "col-sm-12">
<h3>Resighted birds by species and region</h3>
```{r 'resights by species', echo = FALSE, message = FALSE, warning = FALSE}

blitzSpp <-
  read_csv('staticBlitz_processedRecords.csv')

blitzBySpp <-
  blitzSpp %>%
  group_by(spp) %>%
  summarize(n = length(unique(bandNumber))) %>%
  filter(!is.na(spp)) %>%
  spread(spp, n, fill = 0) %>%
  mutate_at(
    c('AMRO', 'BCCH', 'CACH', 'CARW', 'GRCA', 'NOCA', 'NOMO', 'SOSP', 'TUTI'),
    as.integer) %>%
  mutate(region = 'Across regions') %>%
  select(region, AMRO:TUTI)

blitzSpp %>% 
  group_by(region, spp) %>%
  summarize(n = length(unique(bandNumber))) %>%
  filter(!is.na(spp)) %>%
  spread(spp, n, fill = 0) %>%
  mutate_at(
    c('AMRO', 'BCCH', 'CACH', 'CARW', 'GRCA', 'NOCA', 'NOMO', 'SOSP', 'TUTI'),
    as.integer) %>%
  bind_rows(blitzBySpp) %>%
  kable('html') %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c('striped', 'hover', full_width = F)
  )
```
</div>
</div>

<hr>
<br>
<h2>And the participant winners are ...</h2>
<br>
<div class = "row">
<div class = "col-sm-5">
<p><b>Participants with the most resights</b>:</p>
```{r echo = FALSE}
kable(
  blitz %>%
  group_by(siteID) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
    slice(1:10),
  'html') %>%
  kable_styling(full_width = TRUE,
    bootstrap_options = c('striped', 'hover', full_width = F)
    )
```
</div>
<div class = "col-sm-2"></div>

<div class = "col-sm-5">
<p><b>Participants with the most resighted birds</b>:</p>
```{r echo = FALSE}
kable(
  blitz %>%
    select(-date) %>%
    distinct %>%
    group_by(siteID) %>%
    summarize(n = n()) %>%
    arrange(desc(n))  %>%
    slice(1:10),
  'html') %>%
  kable_styling(full_width = TRUE,
    bootstrap_options = c('striped', 'hover')
  )
```
</div>
</div>
<br>
<hr>

<!-- <br> -->
<!-- ```{r echo = FALSE} -->
<!-- blitz %>% -->
<!--   arrange(region, siteID, spp) %>% -->
<!--   kable('html') %>% -->
<!--   kable_styling(full_width = TRUE, -->
<!--                 bootstrap_options = c('striped', 'hover') -->
<!--   ) -->
<!-- ``` -->



